Object-oriented implementations of the MPDATA advection equation solver in~C++,~Python~and~Fortran

### 2013-03-15 paper/doc/paper.tex

%\journal{Computer Physics Communications}
\journal{arXiv. Paper presented at the American Meteorological Society 2013 Annual Meeting in Austin, Texas, USA}

@@

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        Presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers include an implementation of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes exemplify how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        The introduced codes serve as a basis for discussion on the tradeoffs of the programming language choice.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        In case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

@@

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae/adv}
  \end{equation}
  that describe transport of quantity $\psi$ by fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion
    \citep[see][for a recent review of MPDATA-based techniques]{Smolarkiewicz_2006}.

@@

  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation of the $\psi$ and 
    the $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ (Courant number) fields onto a grid - an array of points.
  Among the three considered languages only Fortran is equipped with built-in
    array handling facilities of practical use in high-performance computing.

  Presented C++ implementation of MPDATA is built upon the Blitz++ library. %TODO: cite something
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects, 
    a crucial feature for achieving high-performance in array manipulation
    \citep[][]{Veldhuizen_et_al_1997}.
  Blitz++ is a header-only library\footnote{unless debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code/cpp/listings.hpp}{listing02}{listing03}{1}
  Here, \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} will serve as alias identifiers for the
    2D double precision version of the Blitz++ \prog{Array} 
    container, for the \prog{Range} class representing a sequence of integers numbering 
    array elements, and for the 2D version of the \prog{RectDomain} class representing 
    a vector of \prog{Range} objects defining a 2D slab of an array, respectively.

  Arguably, the power of Blitz++ comes from the possibility to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ,,recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible, and the \prog{auto} 
    return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code/cpp/listings.hpp}{listing03}{listing04}{1}
    in which a call to \prog{blitz::safeToReturn()} function is also included ensuring
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA the NumPy package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python implementation includes the following sequence of \prog{import} statements:
  \codepyt{code/pyt/listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables it to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, the NumPy package is instructed to ignore any floating-point errors, if such an option
    is available in the interpreter (it is not supported in PyPy as of version 1.9).
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

  Fortran features built-in multidimensional array handling mechanisms,
    hence there is no need for using an external package as with C++ and Python.
  Fortran array-handling features are not object-oriented, however.
  In particular, it is not possible to alter the default behaviour through
    inheritance.

  \subsection{Array container sequences}\label{sec:sequence}

  While discretisation in space of the scalar field $\psi(x,y)$ into its $\psi_{[i,j]}$ 
    grid representation requires floating-point array containers (cf. section \ref{sec:array}), 
    discretisation in time requires a container class for storing
    sequences of such arrays, i.e. \{$\psi_{[i,j]}^{[n]}$, $\psi_{[i,j]}^{[n+1]}$\}.
  Similarly the vector field $\vec{C}$ is in fact a \{$C^{[x]}$, $C^{[y]}$\} sequence.

@@

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code/for/listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} and \prog{arrptr\_t} helper types are defined solely for the purpose 
    of overcoming the Fortran's limitation of not allowing to declare arrays of arrays.
  The \prog{ctor()} method initialises the container for a given number of elements \prog{n}.
  The \prog{init()} method initialises the \prog{n}-th element of the container with 
    an newly allocated 2D array spanning indices \prog{i(1)}:\prog{i(2)}, 
    and \prog{j(1)}:\prog{j(2)} in the first, and last dimensions respectively\footnote{When an array
    is passed as function argument in Fortran its base is locally set to unity, regardless of the setting
    at the caller scope, hence the first elements of \prog{i} and \prog{j} are always \prog{i(1)} and \prog{j(1)}.}.
  The \prog{dtor()} method serves as a destructor\footnote{as of time of writing there is no
    open-source Fortran compiler supporting destructor-like FINAL methods defined in the Fortran 2003 standard} 
    and deallocates all the memory allocated
    through previous calls to \prog{init}.
  Access to the arrays stored in the container is provided through the \prog{at} member field.
  Assuming \prog{psi} is an instance of \prog{arrptr\_t}, the \prog{(i,j)} element of the \prog{n}-th array in \prog{psi}
    may be accessed with \prog{psi\%at( n )\%p\%a( i, j )}.

%  TODO: mention why module procedures have to be prefixed with module name in Fortran
  \subsection{Staggered grid}\label{sec:grid}

@@

%TODO: explain why h=(-1,0) is needed with Blitz and Python (in order not to lose memory)

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{Shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}one could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate class}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

@@

  An \prog{ext()} function is defined in all three implementation, in order to simplify 
    coding of array range extensions:
% TODO: works for int 1 and slice
  \codecpp{code/cpp/listings.hpp}{listing06}{listing07}{1}
  \codepyt{code/pyt/listings.py}{listing05}{listing06}{1}
  \codefor{code/for/listings.f}{listing04}{listing05}{1}
  Consequently, a range depicted by $i\pm1/2$ may be expressed in the code as \prog{ext(i, h)}.

%TODO: {discuss how halo region relates to parallel programming}

@@

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}:
  \codecpp{code/cpp/listings.hpp}{listing07}{listing08}{1}
  (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations).

  NumPy uses tuples of slices for addressing multi-dimensional arrays with a
    single object, hence the following definition of function \prog{pi()} suffices to represent $\pi$:
  \codepyt{code/pyt/listings.py}{listing06}{listing07}{1}

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}

%  TODO: templates -> branchless code by design, while F relies on opti

@@

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{description}
    \item[(i)]{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item[(ii)]{allocating and deallocating the required memory,}
    \item[(iii)]{providing access to the solver state, and}
    \item[(iv)]{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{description}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code/cpp/listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with the values of \prog{psi[n]} advected
     within one timestep.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

  The sizes of the arrays representing the two time-levels $\psi^{[n]}$ and $\psi^{[n+1]}$ of $\psi$ 
    are defined by the domain size ({\em nx}$~\times$~{\em ny}) plus the halo region.
  The size of the halo region is an argument of the constructor.

@@

%  TODO: clarify what cycle() does

  Python definition of the \prog{Solver} class follows closely the C++ structure definition:
  \codepyt{code/pyt/listings.py}{listing07}{listing08}{1}
  The key difference stems from the fact that unlike Blitz++, NumPy does not allow an array
    to have arbitrary index base -- in NumPy the first element is always addressed with 0.
  Consequently, while in C++ (and Fortran) the computational domain is chosen to start at (i=0, j=0)
    and hence a part of the halo region to have negative indices, in Python the halo region starts %TODO: footnote o roznych zakresach dla MPI
    at (0,0).
  However, since the whole halo logic is hidden within the solver, such details are not exposed to the
    user.
  The \prog{bcx} and \prog{bcy} boundary-condition specifications are passed to the solver through
    constructor-like \prog{\_\_init\_\_()} method as opposed to template parameters in C++.

  The C++ and Python prototype solver implementations introduced above
    allowed the solvers to operate with any boundary condition
    objects provided the requirements imposed by calls to these objects' methods were fulfilled 
    (at compile-time in case of C++, and at run-time in the case of Python).
  In order to obtain an analogous behaviour with Fortran, it is required to
    define, prior to definition of a solver type, 
    an abstract type with deferred procedures having abstract interfaces 
    \citep[see Table 2.1 in][for a summary of approximate correspondence of OOP 
      nomenclature between Fortran and C++]{Rouson_et_al_2012}:
  \codefor{code/for/listings.f}{listing06}{listing07}{1}
  Having defined the abstract type for boundary-condition objects, 
    a definition of a solver class following closely the C++ and Python counterparts may be provided:
  \codefor{code/for/listings.f}{listing07}{listing08}{1}

  As a side note, it is worth mentioning that the proposed design is readily extendable 
    to handle concurrent computations using domain decomposition.
  The proposed prototype solver is basically ready to be applied 
    in a distributed memory set-up (e.g. using MPI)
    since all necessary communication logic could be placed in the boundary-condition objects
    not requiring any changes to the solver itself.
  Shared-memory parallelisation (e.g. OpenMP) would require minimal changes to the solver structure
    (the \prog{advop()} method would need to be called in a concurrently executed loop),
    leaving the boundary condition- and advection operator-representing logic basically untouched.

  Finally, it has to be underlined that the proposed prototype solver is designed
    just as a minimal example to present the concepts described in the paper.
%  It is by no means an optimal or versatile solution.

@@

  \subsection{Donor-cell formul\ae~(C++)}\label{sec:donor}
  MPDATA is an iterative algorithm in which each iteration takes the form of the 
    so-called donor-cell formula which itself is a first-order advection scheme.
  The set of functions related to the donor-cell scheme is encapsulated within
    the \prog{donorcell} namespace in C++:
  \codecpp{code/cpp/listings.hpp}{listing10}{listing11}{1}

@@

  As mentioned in the previous section, the donor-cell formula
    constitutes an advection scheme, hence we may use it to create
    a \prog{solver\_donorcell} implementation of the abstract \prog{solver} class:
  \codecpp{code/cpp/listings.hpp}{listing15}{listing16}{1}

@@

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step has the form of equation~\ref{eq:donor} with the Courant numbers replaced
    with those defined by the so-called antidiffusive velocities, and with $\psi$ fields from 
    the previous iterations.
  The Courant number field corresponding to the basic MPDATA antidiffusive velocity takes the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:antidiff1}
    \input{formulae/antidiff1}
  \end{equation}
  where
  \begin{equation}\label{eq:antidiff2}
    \input{formulae/antidiff2}
  \end{equation}
  and where for positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:
  \begin{equation}\label{eq:A}
    \input{formulae/A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae/B}
  \end{equation}

@@

  Listings P.13-PP.17 and F.15-F.21 contain the Python and Fortran counterparts to listing C.16-C.22.

@@

  An MPDATA solver may be now constructed with the following definition in C++:
  \codecpp{code/cpp/listings.hpp}{listing23}{listing24}{1}

@@

  The following listing provides an example of how the MPDATA solver
    defined in section~\ref{sec:mpdatasolver} may be used together
    with the cyclic boundary conditions defined in section~\ref{sec:cyclic}.
  In the example a Gaussian signal is advected in a 2D domain 
    defined over a grid of 24$\times$24 cells.
  The program first plots the initial condition, then performs
    the integration for 75 timesteps with three different 
    settings of the number of iterations used in MPDATA.
  The signal shape at the end of each simulation is plotted as well.
  Plotting is done with the help of the gnuplot-iostream library.
  The resultant plot is presented herein as Figure~\ref{fig:mpdata}

  \codecpp{code/cpp/plot.cpp}{listing24}{listing25}{1}

>>>>

\journal{Computer Physics Communications}
%\journal{arXiv. Paper presented at the American Meteorological Society 2013 Annual Meeting in Austin, Texas, USA}

@@

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        Presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers include an implementation of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes exemplify how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        The codes serve also as a basis for discussion on the tradeoffs of the programming language choice.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        In case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

@@

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae/adv}
  \end{equation}
  that describe transport of quantity $\psi$ by fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion.
  All relevant MPDATA formul\ae are given in the text but are presented without
    derivation or detailed discussion.
  For a recent review of MPDATA-based techniques see \citet[][and references therein]{Smolarkiewicz_2006}.

@@

  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of the $\psi$ and 
    the Courant number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.
  Among the three considered languages only Fortran is equipped with built-in
    array handling facilities of practical use in high-performance computing.

  Presented C++ implementation of MPDATA is built upon the Blitz++ 
    library\footnote{Blitz++ is a C++ class library for scientific computing which uses the expression templates technique to achieve high performance, see http://sf.net/projects/blitz/}.
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects.
  Blitz++ is a header-only library\footnote{Blitz++ requires linking with \prog{libblitz} if debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code/cpp/listings.hpp}{listing02}{listing03}{1}
  Here \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} serve as alias identifiers 
    and are introduced in order to shorten the code.

  Arguably, the power of Blitz++ comes from the ability to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ,,recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible.
  Luckily, the \prog{auto} return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code/cpp/listings.hpp}{listing03}{listing04}{1}
  The call to \prog{blitz::safeToReturn()} function is included in order to ensure that
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA the NumPy package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python implementation includes the following sequence of \prog{import} statements:
  \codepyt{code/pyt/listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, to match the settings of C++ and Fortran compilers used herein, the NumPy package is instructed 
    to ignore any floating-point errors (only if such an option
    is available in the interpreter\footnote{\prog{numpy.seterr()} is not supported in PyPy as of version 1.9}).
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

  Fortran features built-in multidimensional array handling mechanisms,
    hence there is no need for using an external package as with C++ and Python.
  Fortran array-handling features are not object-oriented, though.

  \subsection{Array sequence containers}\label{sec:sequence}

  While discretisation in space of the scalar field $\psi(x,y)$ into its $\psi_{[i,j]}$ 
    grid representation requires floating-point array containers (cf. section \ref{sec:array}).
  Discretisation in time requires a container class for storing
    sequences of such arrays, i.e. \{$\psi^{[n]}$, $\psi^{[n+1]}$\}.
  Similarly the components of the vector field $\vec{C}$ are in fact a \{$C^{[x]}$, $C^{[y]}$\} 
    array sequence.

@@

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code/for/listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} type is defined solely for the purpose 
    of overcoming the Fortran's limitation of not allowing to declare arrays of arrays.
  The \prog{arrptr\_t} type is defined as creating an allocatable of its instances instead
    of multi-element pointer of \prog{arr\_t} ensures automatic memory deallocation.
  The \prog{ctor()} method initialises the container for a given number of elements \prog{n}.
  The \prog{init()} method initialises the \prog{n}-th element of the container with 
    an newly allocated 2D array spanning indices \prog{i(1)}:\prog{i(2)}, 
    and \prog{j(1)}:\prog{j(2)} in the first, and last dimensions respectively\footnote{In Fortran, when an array
    is passed as a function argument its base is locally set to unity, regardless of the setting
    at the caller scope.}.
  Access to the arrays stored in the container is provided through the \prog{at} member field.
  Assuming \prog{psi} is an instance of \prog{arrptr\_t}, the \prog{(i,j)} element of the \prog{n}-th array in \prog{psi}
    may be accessed with \prog{psi\%at( n )\%p\%a( i, j )}.

  \subsection{Staggered grid}\label{sec:grid}

@@

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{Shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate class}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

@@

  An \prog{ext()} function is defined in all three implementation, in order to simplify 
    coding of array range extensions:
  \codecpp{code/cpp/listings.hpp}{listing06}{listing07}{1}
  \codepyt{code/pyt/listings.py}{listing05}{listing06}{1}
  \codefor{code/for/listings.f}{listing04}{listing05}{1}
  Consequently, a range depicted by $i\pm1/2$ may be expressed in the code as \prog{ext(i, h)}.
  In all three implementations the \prog{ext()} function accept the second
    argument to be an integer or a ''half'' (cf. section \ref{sec:grid}).

@@

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}
    (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations):
  \codecpp{code/cpp/listings.hpp}{listing07}{listing08}{1}

  NumPy uses tuples of slices for addressing multi-dimensional array with a
    single object.
  Therefore, the following definition of function \prog{pi()} suffices to represent $\pi$:
  \codepyt{code/pyt/listings.py}{listing06}{listing07}{1}

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}
  \prog{span()} is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.11 and F.17--F.20).

  It is worth noting here that the C++ implementation of \prog{pi()} is
    branchless thanks to employment of template specialisation.
  With Fortran one needs to rely on compiler optimisations to eliminate the conditional expression
    within the \prog{pi()} that depends on value of \prog{d} which is always known 
    at compile time.

@@

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{description}
    \item[(i)]{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item[(ii)]{allocating and deallocating the required memory,}
    \item[(iii)]{providing access to the solver state, and}
    \item[(iv)]{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{description}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code/cpp/listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with an updated (advected)
     values of \prog{psi[n]}.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

  The donor-cell and MPDATA schemes both require only the previous state of an advected field
    in order to advance the solution.
  Consequently, memory for two time levels ($\psi^{[n]}$ and $\psi^{[n+1]}$) is allocated in the
    constructor.
  The sizes of the arrays representing the two time-levels of $\psi$ 
    are defined by the domain size ({\em nx}$~\times$~{\em ny}) plus the halo region.
  The size of the halo region is an argument of the constructor.
  The \prog{cycle()} method is used to swap the time levels without copying any data.

@@

  Python definition of the \prog{Solver} class follows closely the C++ structure definition:
  \codepyt{code/pyt/listings.py}{listing07}{listing08}{1}
  The key difference stems from the fact that unlike Blitz++, NumPy does not allow an array
    to have arbitrary index base -- in NumPy the first element is always addressed with 0.
  Consequently, while in C++ (and Fortran) the computational domain is chosen to start at (i=0, j=0)
    and hence a part of the halo region to have negative indices, in Python the halo region starts 
    at (0,0)\footnote{The reason to allow the domain to begin at arbitrary index
    is mainly to ease debugging in case the code would be used in parallel computations
    using domain decomposition where each subdomain could have its own index base corresponding to
    the location within the computational domain}.
  However, since the whole halo logic is hidden within the solver, such details are not exposed to the
    user.
  The \prog{bcx} and \prog{bcy} boundary-condition specifications are passed to the solver through
    constructor-like \prog{\_\_init\_\_()} method as opposed to template parameters in C++.

  The above C++ and Python prototype solvers allow to operate with any boundary condition
    objects provided the requirements imposed by calls to these objects' methods are fulfilled 
    (at compile-time in case of C++, and at run-time in the case of Python).
  In order to obtain an analogous behaviour with Fortran, it is required to
    define, prior to definition of a solver type, 
    an abstract type with deferred procedures having abstract interfaces 
    \citep[sic!, see Table 2.1 in][for a summary of approximate correspondence of OOP 
      nomenclature between Fortran and C++]{Rouson_et_al_2012}:
  \codefor{code/for/listings.f}{listing06}{listing07}{1}
  Having defined the abstract type for boundary-condition objects, 
    a definition of a solver class following closely the C++ and Python counterparts may be provided:
  \codefor{code/for/listings.f}{listing07}{listing08}{1}

@@

  \subsection{Donor-cell formul\ae~(C++)}\label{sec:donor}
  MPDATA is an iterative algorithm in which each iteration takes the form of the 
    so-called donor-cell formula (which itself is a first-order advection scheme).
  The set of functions related to the donor-cell scheme is encapsulated within
    the \prog{donorcell} namespace in C++:
  \codecpp{code/cpp/listings.hpp}{listing10}{listing11}{1}

@@

  As mentioned in the previous section, the donor-cell formula
    constitutes an advection scheme, hence we may use it to create
    a \prog{solver\_donorcell} implementation of the abstract \prog{solver} class: 
  \codecpp{code/cpp/listings.hpp}{listing15}{listing16}{1}
  The above definition is not used in the following MPDATA implementation -- 
    it is given as an example only.

@@

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step is a donor-cell step (eq.~\ref{eq:donor}) with the Courant number
    fields corresponding to the MPDATA antidiffusive velocities of the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:antidiff1}
    \input{formulae/antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where
  \begin{equation}\label{eq:antidiff2}
    \input{formulae/antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:
  \begin{equation}\label{eq:A}
    \input{formulae/A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae/B}
  \end{equation}

@@

  Listings P.13-P.17 and F.15-F.21 contain the Python and Fortran counterparts to listing C.16-C.22.

@@

  An MPDATA solver may be now constructed by inheriting from \prog{solver} class
    with the following definition in C++:
  \codecpp{code/cpp/listings.hpp}{listing23}{listing24}{1}
  The sequence of temporary arrays \prog{tmp} allocated in the constructor
    is used to store the antidiffusive velocities from the present and optionally 
    previous timestep (if using more than two iterations).

  The \prog{advop()} method controlls the MPDATA iterations within one timestep.
  The first (step = 0) iteration of MPDATA is an unmodified donor-cell step (see listing C.15).
  Subsequent iterations begin with calculation of the antidiffusive Courant fields using 
    formula \ref{eq:antidiff1}.
  In order to calculate values spanning an (i\mhlf~... i+\phlf) range using a formula
    for $C_{[i+\phlf,\ldots]}$ only, the formula is evaluated using extended
    index ranges \prog{im} and \prog{jm}.
  In the second (step=1) iteration the uncorrected Courant field (\prog{C\_unco}) points
    to the original \prog{C} field, and the antidiffusive Courant field is written into
    \prog{C\_corr} which points to \prog{tmp[1]}.
  In the third (step=2) iteration \prog{C\_unco} points to \prog{tmp[1]} while \prog{C\_corr}
    points to \prog{tmp[0]}.
  In subsequent iterations \prog{tmp[0]} and \prog{tmp[1]} are alternately swapped.

  Listings P.18 and F.22 contain the Python and Fortran counterparts to listing C.23.

@@

  The following listing provides an example of how the MPDATA solver
    defined in section~\ref{sec:mpdatasolver} may be used together
    with the cyclic boundary conditions defined in section~\ref{sec:cyclic}.
  In the example a Gaussian signal is advected in a 2D domain 
    defined over a grid of 24$\times$24 cells.
  The program first plots the initial condition, then performs
    the integration for 75 timesteps with three different 
    settings of the number of iterations used in MPDATA.
  The velocity field is constant in time and space (although it is
    not assumed in the presented implementations).
  The signal shape at the end of each simulation is plotted as well.
  Plotting is done with the help of the gnuplot-iostream library.

  The resultant plot is presented herein as Figure~\ref{fig:mpdata}.
  The top panel depicts the initial condition. 
  The three other panels show a snapshot of the field after 75 timesteps.
  The donor-cell solution is characterised by strongest numerical diffusion
    resulting in significant drop in the signal amplitude.
  The signals advected using MPDATA show smaller numerical diffusion.
